# Working with R - Answers {.unnumbered}

```{r setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r, echo = FALSE, purl = FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)

options(purl = FALSE)

knitr::opts_chunk$set(purl = FALSE)
```

```{r, include = FALSE}
knitr::opts_hooks$set(eval = function(opt) {
  if (any(opt$exercise))
    opt$eval <- opt$include <- FALSE
  
  opt
})


static <- TRUE

options(width = 100)

```

```{r packages, include = FALSE}
library(kableExtra)
library(knitr)
```

```{r load_data, context="data", include=FALSE}
library(survival)
```

## Preface {data-progressive="FALSE"}

Open RStudio to do the practicals. Note that tasks with `*` are optional.

This answer file follows the same structure as the practical. Use it to check both your code and your understanding. Try to compare your own script with the solutions before copying anything over directly.

By the end of this practical, you should be able to:

- use `R` as a calculator
- assign results to variables
- create simple vectors
- load a package and read help pages
- save, load and remove objects from your workspace

### R packages

In this practical, a number of R packages are used. The packages used (with versions that were used to generate the solutions) are:

-   `survival` (version: `r packageVersion("survival")`)

*`r R.version.string`*

### Dataset

For this practical, we will use the **heart** and **retinopathy** data sets from the `survival` package. More details about the data sets can be found in:

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/heart.html

https://stat.ethz.ch/R-manual/R-devel/library/survival/html/retinopathy.html

We will learn more about the format of data sets (or `data.frame`s) later.

::: callout-tip

Try to solve the tasks on your own first. If you get stuck, look at the hints and try to understand the code in the hints. If you still do not understand, ask for help. 
When you work on the the tasks, place your code in a script. This way you can save your work and rerun it later. Also you build good habits right from the start.

As you work through the tasks, occasionally look at the Environment pane in RStudio. It is a useful way to connect the code you run to the objects that now exist in your session.
:::

## Basic calculations

### Expressions {.tabset .tabset-fade .tabset-pills}

::: {.panel-tabset .nav-pills}
#### Task 1

Use R as a calculator to calculate the following values:

-   $45 - 2 \times 3$
-   $(45 - 2) \times 3$
-   $17 ^4$,
-   $2 ^  (1/3)$,


#### *Hint 1*

Enter the calculations in an R-script and then run the lines using {{< kbd win=Ctrl+Enter mac=Cmd+Enter >}}.

```{r eval=FALSE, include=TRUE}
45-2*3
(45-2)*3
#etc
```


#### Solution 1

```{r save-expres, solution = TRUE, eval=F, include=TRUE}

45-2 * 3
(45-2) * 3
17^4
2^(1/3)

```

Notice that each expression is evaluated immediately. At this stage, you are not storing anything yet; you are simply using `R` to calculate results.

:::


### Assignment {.tabset .tabset-fade .tabset-pills}
::: panel-tabset

#### Task 2

Now use assignment to do the same calculation. Assign names `a`, `b` and `c` to each of the three values involved, then do the calculation while assigning the name `d` to the result.

#### Hint 2

Execute the following code.

```{r}
a <- 45
b <- 2
c <- 3
d <- (a - b) * c 
```

`a`, `b`, `c`, and `d` are variables. To see their values, you can just type the variable name (e.g. `a`) and hit {{< kbd Ctrl-Enter >}} or use the command `print(a)`.

#### Solution 2 { .unnumbered}

```{r sol2, solution = TRUE, eval=FALSE}
a <- 45
b <- 2
c <- 3
d <- (a - b) * c 
print(a)
print(b)
print(c)
print(d)
d
```

Here the important thing is not only the final value of `d`, but also the fact that the intermediate values are stored and can be reused later.
:::


::: panel-tabset

R can do calculations on whole vectors at once. 

#### Task 3

Combine the two values 3 and 4 into a vector `v1` using the `c` function. Assign the name `v1` to the vector. Then calculate the square of each value in the vector. Create a vector `v2` consisting of the numbers 1 and 2 and calculate the sum of `v1` and `v2`.

#### Hint 3

Execute the following code.

```{r}
v1 <- c(3, 4)
# now compute the square of each value
# create v2 in the same way as v1 and calculate the sum
```




#### Solution 3 {.unnumbered}

```{r sol3, solution = TRUE, eval=FALSE }
v1 <- c(3, 4) # the c(.) function combines the arguments 3 and 4 into a vector
v1^2
v2 <- c(1, 2)
v1 + v2
```

This task introduces a key idea in `R`: many operations work on whole vectors at once.

:::


::: panel-tabset

A type of vector that is used often is simply a sequence of numbers

#### Task 4

Create a vector of numbers using the function call `seq(1, 9)`. Check that `1:9` does the same thing. What do `seq(1, 9, by=2)` and `9:1` do?

#### Hint 4


```{r}
# Just type in the given code fragments and see what they do.
```

#### Solution 4 {.unnumbered}

```{r sol4, solution = TRUE, eval=FALSE}
seq(1, 9)
1:9
seq(1, 9, by=2)
9:1
```

The point here is to see that `R` has several convenient ways to generate sequences.

:::

::: callout-note
## Milestone

At this point, your script should contain a few simple calculations, some assignments and at least one vector. If you rerun the script from the top, it should work without errors.
:::

## Loading packages and using the help function

### Loading packages {.tabset .tabset-fade .tabset-pills}

We will use data sets from the `survival` package. It is important to know how to load packages as they contain most of the functionality of R. 


::: {.panel-tabset .nav-pills}
#### Task 1

 Load the `survival` package using the `library` function.

#### *Hint 1*

Use the `library(...)` function. There should be no need to install the package as it is already installed with `R`.

#### Solution 1 {.unnumbered}

```{r, solution = TRUE, eval=FALSE}
library(survival)
# library("survival") is an alternative in this case 
# require(survival) is another alternative which does not give an error if the package cannot be found
```

The key habit here is to load packages explicitly at the top of your script.

:::

### Getting help {.tabset .tabset-fade .tabset-pills}

::: {.panel-tabset .nav-pills}
#### Task 2

 Ask for the documentation of the `heart` data set, using the `help` function. Then do the same for the `retinopathy` data set using `?`.

#### *Hint 2*

Use the `help(...)` function and `?`. 

#### Solution 2 {.unnumbered}

```{r help, solution = TRUE, eval=FALSE}
help("heart")
?retinopathy
```

Both forms open documentation. It is useful to become comfortable with both.

:::

::: {.panel-tabset .nav-pills}
#### Task 3

Also ask for help on the `c`, `ls`,  `rm` and `seq` functions. Check if you can also obtain the documentation for the `c` function if you use `??combine` 

#### *Hint 3*

Use the `help(...)` function, and `?` or `??`. 

#### Solution 3 {.unnumbered}

```{r help2, solution = TRUE, eval=FALSE}
?c # or use help("...")
?ls
?rm
?seq
??combine
# or use help.search('...')
```

The main lesson is that you do not need to memorise everything. A large part of learning `R` is learning how to find help efficiently.

:::

::: callout-note
## Milestone

You have now used three important habits of good `R` work:

- loading packages explicitly
- checking documentation when you do not remember syntax
- exploring functions before using them
:::

## Importing and Saving Data

### Save your work {.tabset .tabset-fade .tabset-pills}

It is important to save your work. You can save the whole workspace using the function `save.image`. If you want to save only specific objects, you can use the function `save`.

::: {.panel-tabset .nav-pills}
#### Task 1

Save the vectors `numbers <- c(34, 24, 19, 23, 16)`, `numbers_2 <- c(1:200)` and `treatment <- c("yes", "yes", "no", "no", "no", "yes")`. Use the name `new_vectors`.

#### *Hint 1*

Use the function `save(...)`. Note that you need to set the working directory.

#### Solution 1

```{r save-solution, solution = TRUE, eval=FALSE}
numbers <- c(34, 24, 19, 23, 16)
numbers_2 <- c(1:200)
treatment <- c("yes", "yes", "no", "no", "no", "yes")
save(numbers, numbers_2, treatment, file = "new_vectors.RData")
```

This saves several objects together in a single `.RData` file.
:::

::: {.panel-tabset .nav-pills}
#### Task 2

Save the vectors `events <- heart$event` and `eyes <- retinopathy$eye`. Use the name `vectors_survival`.

#### *Hint 2*

Use the function `save(...)`. Note that you need to set the working directory.

#### Solution 2

```{r save2-solution, solution = TRUE, eval=FALSE}
events <- heart$event
eyes <- retinopathy$eye
save(events, eyes, file = "vectors_survival.RData")
```

This is similar to the previous task, but now the saved objects come from real data sets rather than values you typed yourself.
:::

### Load your work {.tabset .tabset-fade .tabset-pills}

Let's continue working on the data sets by loading our results.

::: {.panel-tabset .nav-pills}
#### Task

Load the file `new_vectors`.

#### *Hint*

Use the function `load(...)`.

#### Solution

```{r load-solution, solution = TRUE, eval=F}
load("new_vectors.RData")
```

After loading, the saved object names reappear directly in the workspace.
:::

At this point it is interesting to check the objects that are in the workspace. To do so look at the Environment tab in the upper-right hand pane. You can also use the `ls()` function to list all objects in the workspace.

Try to answer this question before moving on: which objects came from your earlier code, and which appeared only after loading the saved file?

### Remove your work {.tabset .tabset-fade .tabset-pills}

Remove unnecessary objects.

::: {.panel-tabset .nav-pills}
#### Task 1

Remove the vectors `numbers`, `numbers_2` and `treatment`.

#### *Hint 1*

Use the function `rm(...)`.

#### Solution 1

```{r rm-solution, solution = TRUE, eval=F}
rm(numbers, numbers_2, treatment)
```

`rm()` removes objects from the current workspace, but it does not delete your script file or the saved `.RData` file on disk.
:::

::: {.panel-tabset .nav-pills}
#### Task 2 \*

Remove the vectors `events` and `eyes`.

#### *Hint 2* \*

Use the function `rm(...)`.

#### Solution 2 \*

```{r, eval=F}
rm(events, eyes)
```

Again, this only removes the objects from memory in the current session.
:::


::: callout-note
## Final check

Before you finish this practical:

- inspect the History tab in the upper-right pane
- if you wrote a script to work on this practical, check whether your script runs from top to bottom without errors. Also do dot forget to save it to disk!
- make sure you understand which objects are still present in the workspace and why
:::
