# Indexing and Subsetting {.unnumbered}

**Authors: Karl Brand, Elizabeth Ribble and Sten Willemsen**

## Introduction

Subsetting is one of the practical skills in `R` that you end up using all the time. In real analyses, we almost never want the whole data set in exactly the form it arrived. We want to select the adult patients, exclude the rows with missing values, or to use only a few variables for the next step. We want one element from a list, one column from a table, or one slice of a matrix. That is what this chapter is about.

Once this starts to feel natural, a lot of later work becomes easier. Data cleaning, plotting and modelling (things we will discuss in the next couple of days) all depend on being able to pull out the part you actually need. We will begin with vectors and then move to matrices, arrays, lists and `data.frame`s, because the basic idea stays the same even though the objects become richer.

## A quick cheat sheet

Before diving into the details, it helps to keep the most common patterns in one place:

| Pattern | Meaning |
|---|---|
| `x[i]` | select element(s) from a vector by position |
| `x[name]` | select element(s) from a named vector by name |
| `x[condition]` | select element(s) where the condition is `TRUE` |
| `m[row, col]` | select from a matrix |
| `a[i, j, k]` | select from an array |
| `lst[i]` | select element(s) from a list and keep the result as a list |
| `lst[[i]]` | extract a single list element itself |
| `lst$name` | extract a named list element |
| `df[row, col]` | select rows and columns from a `data.frame` |
| `df["col"]` | select one column as a `data.frame` |
| `df[["col"]]` or `df$col` | select one column as a vector |

::: callout-warning
### A common source of confusion

Two pieces of code can look almost the same but return different kinds of objects:

- `df["age"]` returns a one-column `data.frame`
- `df[["age"]]` returns the vector itself
- `df[, "age"]` often returns a vector
- `df[, "age", drop = FALSE]` keeps the result as a `data.frame`

When in doubt, check with `class(...)` or `str(...)`.
:::

## Indexing a vector

There are three ways in which we can make a selection:

-   Using numeric values, to make selections based on the position in a vector
-   Using character values, to select values based on their name
-   Using logical values (TRUE / FALSE), to make selections based on a condition

### Using numeric values

The easiest way of selecting elements in a vector is by specifying positions, or **indices** `i`, using square brackets, that is `object[i]`.

```{r}
v <- c(11, 13, 17, 15, 9)
v[2]
```

We can also do this for multiple elements at the same time:

```{r}
v[c(2, 3)]
i <- c(1, 2)
v[i]

```

The vector of indices does not have to be sorted and indices may be duplicated. For example:

```{r}
v[c(3, 2, 2)]
```

When we use a vector with negative integers we will select all observations *except* those on the specified positions:

```{r}
v[-2]

```

Note that positive and negative indices cannot be combined.

### Using character values

When the elements in a vector all have a name we can use these names to select the elements.

```{r}
bp_with_name <- c(sys=135, dia=85)
bp_with_name['dia']
```

We can also give names to the vector by using the `names` function: Here we make use of the `names` attribute:

```{r}
ages <- c(12, 3, 45)
names(ages) <- c("Kim" , 'Arthur', 'Mark')
ages
str(ages)
ages["Mark"]
names(ages)
```

Note that we can also change the names of a `data.frame` in the same way we would do so for a `list` using the `names` attribute:

```{r}
mygenes <- data.frame(samp1 = c(33, 22, 12), 
                      samp2 = c(44, 111, 13), 
                      samp3 = c(33, 53, 65))
names(mygenes) <- c("samp10", "samp20", "samp30")
mygenes
## but let's change it back...
names(mygenes) <- c("samp1", "samp2", "samp3")
```

Note that the `colnames` function also performs the same job for data frames.

### Using logical values

The third way to select elements from a vector is by using a vector of logical values (i.e.. TRUE/FALSE values) between the square brackets. In this way we select all values for which the value between the brackets is TRUE.

```{r}
some_values<- c('foo', 'bar', 'baz')
some_values[c(TRUE, FALSE, TRUE)]
```

This way of selecting is especially useful when we use some comparison (using `<`, `<=`, `!=` , etc.) or condition to select variables.

```{r}
ages <- c(55, 78, 92, 44)
sex <- factor(c('Male', 'Female', 'Male', 'Female'))
ages[ages > 65]
# When we use a factor as filter we may compare to a character value
ages[sex == 'Female'] 
```

To understand why this works look at when the condition between the brackets is TRUE.

```{r}
ages > 65
which(ages > 65)  # gives the TRUE indices
```

Note that when we apply filtering on a factor variable the possible levels are not changed:

```{r}
levels(sex[sex == 'Female'] )
```

## Making selections in a matrix or array

Making selections on a matrix works more or less the same as for vectors. However, because a matrix has two dimensions, we need two indices: the first for the rows and the second for the columns.

```{r}
m <- matrix(c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12), ncol = 3)
m[1, ] # first row
m[ , 2] # second column
m[c(2, 4), 1:2] # combination of rows and colums
m[ , -c(1, 3)]   # negative indices 
m[m[ , 1]>=3, ] #logical indices
```

Note that when a single row or column is selected, the object is converted to a vector. This is a frequent source of errors. If you want to prevent this, you can use `drop = FALSE`.

```{r}
m[1, , drop=FALSE]
```

We can also use names but a matrix can have rownames as well as colnames:

```{r}
rownames(m) <- LETTERS[1:4]
m
m["A", ]
```

As shown above, we usually use two indices to select elements in a matrix. However, it is also possible to use a single index. In this case, `R` treats the matrix as a vector by stacking the columns on top of each other.

```{r}
m[6]

```

::: callout-advanced

## Using a single index

There is a special way of using a matrix with two columns to select individual elements from another matrix based on two-dimensional coordinates.

```{r}
a <- matrix(c(2,3,3,2), ncol = 2)
a
m[a]
```

## Arrays

Indexing for arrays is similar as for matrices.

```{r}
a <- array(c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),dim = c(2,3,4))
dimnames(a) <- list(NULL, c('A', 'B', 'C'), c('a', 'b', 'c', 'd'))
a
colnames(a) # still 2nd dimension
a[1, 2, ] # 3-dim array requires 3 positions between the brackets
a[,,'a']

```

We can use `drop=FALSE` here as well to prevent R from converting the result to a lower dimension.

```{r}
a[c(TRUE, FALSE), , ]
a[c(TRUE, FALSE), , , drop=FALSE]
```


:::

## Making selections in a list

To make selections in a list, we can use single square brackets, double square brackets (`[[` and `]]`) and the dollar sign. The use of single square brackets works in the same way as it does for vectors.

```{r}
mylist <- list(
  foo=c(1,2, 3),
  bar=c('a', 'b'),
  baz=list(TRUE, c(2, 4))
  )
mylist[c(1,3)]
mylist[c(1,2,3)==2]
mylist['bar']
```

We can also use double square brackets and the dollar sign for a `list`. There are two important differences between using single and double square brackets:

1. Using double square brackets only allows us to select a single element from the `list`.
2. The result of a selection with double brackets is the element itself, while if we make the selection with single brackets the result is a `list` consisting of the selected elements.

```{r}
mylist[[1]]
```

```{r}
mode(mylist[[1]]) # a vector
mode(mylist[1])   # a list with with a numeric vector as its single element   
```

There is another way to select a variable from a `list` which is by using the dollar sign ('\$'). This is an alternative to using double square brackets in combination with a name. When we use this the result is always a vector.

```{r}
mylist$foo      # results is a vector
```

Instead of using the dollar sign we can use double square brackets. We now need to put the name between quotes like for single square brackets. We can also use the position of the variable using these double square brackets.

```{r}
mydata <- data.frame(id=c(1, 2, 3, 4, 5),
                     sex=factor(c('M', 'F', 'M', 'F', 'F')),
                     weight=c(77, 44, 56, 88, 49),
                     treatm=c('A', 'A', 'B', 'B', 'A'))
mydata[['treatm']]
mydata[[1]]
```

It is not possible to select more than one element using double brackets. The result will be the extracted element itself, rather than a `data.frame` or `list`.

## Selecting observations and variables in a `data.frame`

Making selections in a `data.frame` sometimes behaves like making selections in a matrix, but sometimes it behaves like selecting from a list. When we use a double index, selecting observations and variables in a `data.frame` works much like in a matrix. Because a `data.frame` is two-dimensional, we can place two indices between the square brackets. The first corresponds to the observations (rows) and the second corresponds to the variables (columns). For example, we can select the first two observations from the third variable in the `data.frame` using:

```{r}
mydata <- data.frame(id=c(1, 2, 3, 4, 5),
                     sex=factor(c('M', 'F', 'M', 'F', 'F')),
                     weight=c(77, 44, 56, 88, 49))
mydata[c(1, 2), 3]
```

When the first or second position is left blank all rows or columns are selected. For example:

```{r}
mydata[, c(2, 3)]       # sex and weight (2nd and 3rd variable) for all patients
mydata[c(1,2), ]  # all variables for the first two patients
mydata[c(-3,-4), 'sex'] # Negative numbers and names can also be used
```

We have to be careful when we want to select a single variable from a `data.frame`, as we do above. The result will now no longer be a `data.frame` but it is transformed to a `vector`. When we want to prevent this we can use `drop=FALSE`, as follows:

```{r}
mydata[ , 3]
mydata[ , 3, drop=FALSE]       
```

Note that if we select a single row the result is still a `data.frame`. Conversion to a vector would not be possible in this case as a row of a `data.frame` may contain multiple data types.
```{r}
mydata[2 , ]

```

We can also use a single index between the square brackets. This works as if the `data.frame` were a list of variables, that is, its columns.

```{r}
mydata[1]
mydata[[1]] # double brackets converts to vector
```

A `data.frame` has `row.names` (note the dot) as well as variable names we can use for selection. Let's look at an example where row names are gene symbols and column names are sample IDs:

```{r}
mygenes <- data.frame(samp1 = c(33, 22, 12), 
                      samp2 = c(44, 111, 13), 
                      samp3 = c(33, 53, 65))
row.names(mygenes) <- c("CRP", "BRCA1", "HOXA")
names(mygenes)
mygenes
mygenes["CRP", ]
mygenes[, "samp1"]
mygenes[, c("samp1", "samp3")]
mygenes["HOXA", "samp2"]
```

### Other useful functions

Besides square brackets (`[`, `[[`), other useful functions exist for selecting data: `duplicated`, `match`, `%in%`, `grep`, `is.na` and `$`.

To select e.g. rows that are not duplicated:

```{r}
mm <- matrix(c(1, 1, 2, 2), nrow = 4, byrow = TRUE)


mm
mm[!duplicated(mm), ]
```

The above can also be done with **unique**, but the use of **duplicated** might be more appropriate in more complex situations:

```{r}
unique(mm)
```

Calling `match` returns the position of the first match of its first argument in the second argument:

```{r}
match(c("a", "b"), c("a", "c", "a", "b", "a", "b"))
```

whereas `\%in\%` tells you whether the elements of the first argument appear in the second argument:

```{r}
c("a", "b", "d") %in% c("a", "c", "a", "b", "a", "b")
```

Recall our data frame mygenes:

```{r}
mygenes
is.data.frame(mygenes)
```

Note that since `mygenes` is a data frame, it is therefore also an array, which means we can select by the name of the elements in the array:

```{r}
mygenes[match(c("samp1", "samp3"), colnames(mygenes))]
mygenes[colnames(mygenes) %in% c("samp1", "samp4")]
```

However, in this case we could just use the names...

```{r}
mygenes[c("samp1", "samp3")]
```

But this gives an error:

```{r}
#| eval: false
mygenes[c("samp1", "samp30")] ## not run
```

where this does not:

```{r}
mygenes[colnames(mygenes) %in% c("samp1", "samp30")]
```

We can also use functions like `grep` to search for the names we are interested in:

```{r}
mygenes[grep(2, names(mygenes))]
mygenes[grep("A", row.names(mygenes)), ]

```

If we want to find or exclude data with missing values, we can use `is.na`:

```{r}
z <- c(1:4, NA, 5:10)
z
is.na(z)
which(is.na(z))
z[!is.na(z)]
```

## Check yourself

Before moving on, make sure you can answer the following:

1. What is the difference between `x[2]` and `x[c(2, 3)]`?
2. What is the difference between `df["age"]`, `df[["age"]]` and `df[, "age"]`?
3. Why is `drop = FALSE` sometimes important?
4. When would you use a logical condition inside brackets?
