# Basic Markdown {.unnumbered}

### Markdown

Quarto files consist of code and markdown. We have discussed how to code in the previous chapters and will discuss markdown here. Markdown is a specific kind of markup language. What we mean by that is that the text in the document also indicates how it should be formatted in the result (For example if we surround a word by double asterikses (like so \*\*bold\*\*) it will **bold** in the resulting output. Below we will discuss the different types of markup that you can use.

#### Basic Formatting

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(tibble)
library(knitr)
tt <- tribble(~Syntax, ~Output,
              "\\*italics\\*", "*italics*",
              "\\*\\*bold\\*\\*", "**bold**",
              "\\*\\*\\*bold italics\\*\\*\\*", "***bold italics***",
              "subscript\\~2\\~", "subscript~2~",
              "superscript\\^2\\^", "superscript^2^",
              "\\~\\~crossed out\\~\\~", "~~crossed out~~",
              "\\`code\\`", "`code`",
              )
kable(tt)

```

#### Headings

+--------------------+-----------------------------------+
| Syntax             | Output                            |
+====================+===================================+
| ```                | # Header 1 {.heading-output}      |
| # Header 1         |                                   |
| ```                |                                   |
+--------------------+-----------------------------------+
| ```                | ## Header 2 {.heading-output}     |
| ## Header 2        |                                   |
| ```                |                                   |
+--------------------+-----------------------------------+
| ```                | ### Header 3 {.heading-output}    |
| ### Header 3       |                                   |
| ```                |                                   |
+--------------------+-----------------------------------+
| ```                | #### Header 4 {.heading-output}   |
| #### Header 4      |                                   |
| ```                |                                   |
+--------------------+-----------------------------------+
| ```                | ##### Header 5 {.heading-output}  |
| ##### Header 5     |                                   |
| ```                |                                   |
+--------------------+-----------------------------------+
| ```                | ###### Header 6 {.heading-output} |
| ###### Header 6    |                                   |
| ```                |                                   |
+--------------------+-----------------------------------+

```{=html}
<style type="text/css">
.heading-output {
  border-bottom: none;
  margin-top: 0;
  margin-bottom: 0;
}
</style>
```
#### Links

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(tibble)
library(knitr)
tt <- tribble(~Syntax, ~Output,
              "  &lt;https://www.erasmusmc.nl&gt; ", " <https://www.erasmusmc.nl>",
              "  \\[EMC\\](https://www.erasmusmc.nl) ", " [EMC](https://www.erasmusmc.nl)"
              )
kable(tt)

```

#### Images

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(tibble)
library(knitr)
tt <- tribble(~Syntax, ~Output,
              "  !\\[Caption\\](Logo.svg) ", " ![Caption](assets/Logo.svg) ",
              '  !\\[Caption\\](Logo.svg) "hovertext"', ' ![Caption](assets/Logo.svg "hovertext") ',
              '  !\\[Caption\\](Logo.svg){fig-alt="Alt text"} ', ' ![Caption](assets/Logo.svg){fig-alt="Alt text"} '
              )
kable(tt)

```

It is also possible to include images in links (as in `[![Caption](Logo.svg)](https://www.erasmusmc.nl/)`).

#### Lists

+------------------------------------+---------------------------------+
| Markdown Syntax                    | Output                          |
+====================================+=================================+
| ```                                | -   item 1                      |
| * item 1                           |                                 |
| * item 2                           | -   item 2                      |
|     + sub-item 1                   |                                 |
|     + sub-item 2                   |     -   sub-item 1              |
|         - sub-sub-item 1           |                                 |
| ```                                |     -   sub-item 2              |
|                                    |                                 |
|                                    |         -   sub-sub-item 1      |
+------------------------------------+---------------------------------+
| ```                                | -   item 2                      |
| *   item 2                         |                                 |
|                                    |     Part of item (Extra spaces) |
|     Part of item (Extra spaces)    |                                 |
| ```                                |                                 |
+------------------------------------+---------------------------------+
| ```                                | 1.  ordered list                |
| 1. ordered list                    |                                 |
| 2. item 2                          | 2.  item 2                      |
|     i) sub-item 1                  |                                 |
|     ii) sub-item 2                 |     i)  sub-item 1              |
|          A.  etc.                  |     ii) sub-item 1              |
| ```                                |                                 |
|                                    |     ```                         |
|                                    |     A.  etc.                    |
|                                    |     ```                         |
+------------------------------------+---------------------------------+

#### Tables

There are multiple ways to make tables

The easiest way is to use a so called pipe-table like this:

```         
 | foo|  bar  | Left align | Right Align | Center |
 |----|-------|:-----------|------------:|:------:|
 | 1  |    2  | 1  |    2  | 1|  
 | 3  |    4  |3  |    4  | 4 |
 
 : an optional caption
```

The result is:

| foo | bar | Left align | Right Align | Center |
|-----|-----|:-----------|------------:|:------:|
| 1   | 2   | 1          |           2 |   1    |
| 3   | 4   | 3          |           4 |   4    |

: an optional caption

These tables use the pipe character (|) to separate columns and they use dashes (-) to separate the header from the rest of the table. As shown in the example the header seperator can also include colons (:) to indicate left, right or center alignment of the columns.

The first and last pipe character at each row are optional, the ones separating the columns are not although they do not have to be aligned. The header cannot be omitted in a pipe table. The relative width is determined by the number of dashes. An alternative way to specify the relative column widths is by including the attribute after the caption, like this:

``` markdown
: an optional caption {tbl-colwidths="[40,15,15,15,15]"} 
```

Sometimes it is easiest to directly make the tables from R code.

#### Source code

You can use delimited blocks with three backticks to use source code formatting. The syntax:

\`\`\`\`

foo \<- bar

\`\`\`\`

Becomes:

```         
foo <- bar
```

It is possible to add an R to indicate the R language is used and apply apropriate highlighting:

```` markdown
``` R
foo <- bar
```
````

Becomes:

``` r
foo <- bar
```

::: callout-advanced

#### Math

You can use **TeX**-like syntax to format mathematical expressions. This creates good looking formulas and equations in your documents. Note that TeX is a comprehensive typesetting system with extensive capabilities. We cannot posibly cover all of it in this section. Therefore we only provide the bare essentials. For advanced LaTeX features, consult the resources listed at the end of this section.
Inline formulas (i.e. formulas that appear in the running text) are delimited by single \$s and display math (i.e. formulas that appear on a line of their own) uses double \$\$s.

```{r echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(tibble)
library(knitr)
tt <- tribble(~Syntax, ~Output,
              "  \\$a^2=b^2+c^2\\$", " $a^2=b^2+c^2$ ",
              '  \\$\\$a^2=b^2+c^2\\$\\$', ' $$a^2=b^2+c^2$$ '
              )
kable(tt)

```

Geek characters can be created by writing their names preceded by a backslash. For example `\alpha` produces $\alpha$, `\beta` produces $\beta$ and so on. Capitals can be generated by writing the first letter of their name in uppercase  (So `\Omega` for $\Omega$). 

Superscripts can be created using the caret `^` and subscripts using the underscore `_`. For example `x_i` produces $x_i$ and `x^{n}` produces $x^{n}$.
Fractions can be created using `\frac{numerator}{denominator}`. For example `\frac{a}{b}` produces $\frac{a}{b}$. Use `\sqrt{expression}` to create square roots. For example `\sqrt{x}` produces $\sqrt{x}$. Sums and integrals can be created using `\sum` and `\int` respectively. For example `\int_{a}^{b} f(x) dx` produces $\int_{a}^{b} f(x) dx$.

When you want to learn more about TeX, a good resource is the [The LaTeX Companion](https://www.amazon.nl/-/en/Mittelbach-Frank/dp/0201362996).

:::
