# Code Blocks {.unnumbered}

## Introduction

The real power of Quarto shows up when text and code start living in the same document. That is the step from "a nicely formatted note" to "a report that can actually run the analysis it describes." Once you have that, you can write an explanation, insert a code block, render the document, and have the output appear in the right place without any copying and pasting.

So this section is about the mechanics that make that possible. We will look at code chunks, what rendering actually does, how chunk options change what runs and what gets shown, and how short bits of inline code can keep the prose tied to the analysis.

## Code blocks

When we include a code block, also called a **chunk**, we start by writing three backticks followed by the name of the language in braces. In this course, that language will almost always be `r`. The chunk is then closed with another line of three backticks.

The basic structure looks like this:

```` markdown
```{r}
1 + 1
```
````

When the document is rendered, Quarto will run this code and insert the result into the final output unless you tell it not to.

Optionally, you can also include a chunk name after the language. This helps when errors occur, because the chunk name can appear in error messages. In addition, figures created inside the chunk may use that name in their file names.

```` markdown
```{r examplename}
summary(c(1, 2, 3))
```
````

By default, when the document is rendered, the code in the chunks will be executed and both the source code and the output will be included in the document. A document can contain many code chunks, and they are executed in order. Variables defined in one chunk can usually be used in later chunks, just as if the code were part of a single `R` script.

## Chunk Options

It is possible to modify how code blocks behave. This is done with **chunk options**. These options let you control whether code is executed, whether code is shown, whether messages appear, and so on.

One way is to include the options as special comments at the top of the chunk:


```` markdown
```{r}
#| echo: false

# some code

```
````


Alternatively, we can include options directly in the chunk header:

```` markdown
```{r, echo=FALSE}

# some code

```
````

Some options that are often used are:

eval

:   when TRUE (the default), the code is evaluated. When FALSE, it is not.

echo

:   when TRUE (default), the source code is displayed in the output document. When it is FALSE, it is not.

results

:   specifies how the results are displayed. When it is `markup`, the result is displayed as a code block. When it is `asis`, it will not be modified. This is convenient when an `R` function already formats its output in markdown or HTML. When it is `hide` or `FALSE`, the results are not shown.

warning

:   when FALSE, all warnings will be suppressed

message

:   when FALSE, all messages (printed by the `message` function) will be suppressed

include

:   when TRUE, the output is shown in the document. When FALSE, all output is suppressed, but the code is still executed.

file

:   the code content is read from an external file named in the option

child

:   the content of an external file is inserted in place

::: callout-note
### A practical way to think about chunk options

Chunk options answer questions such as:

- Should this code run?
- Should the reader see the code?
- Should the reader see the output?
- Should warnings and messages be shown?

These decisions help you balance transparency and readability in the final report.
:::

## Inline code

Besides complete code blocks, it is also possible to include short bits of `R` code in the running text. The code should be included between `` `r `` and `` ` ``, for example `` `r nrow(mtcars)` ``.

This is useful when you want a sentence in the report to update automatically, for example:

```` markdown
The data set contains `r nrow(mtcars)` observations.
````

When you use inline code, make sure the object you refer to already exists earlier in the document. Otherwise the document will fail to render.

Inline code is often one of the nicest features of Quarto because it lets your text and your results stay synchronized.

## YAML header

Each Quarto document starts with a header that specifies things such as the title, author, date and output format. This header is written in YAML and appears at the top of the file. For example:

``` markdown
---
title: "My document"
author: "Sten Willemsen"
date: "21 May 2023"
format: html
---
```

Besides HTML we can also create Word documents and PDF documents. The PDF option usually requires a TeX installation.

## What to remember

Markdown gives structure to the text of the report, while code blocks allow the analysis itself to be executed inside the document. Together, these are the two core ingredients of Quarto.

## Check yourself

Before moving on, make sure you can answer the following:

1. What is the basic structure of a Quarto code chunk?
2. What is the difference between `echo: false` and `include: false`?
3. Why can inline code be useful in a report?
4. What is the purpose of the YAML header?
