# Shiny {.unnumbered}

## Introduction

Up to this point, most of our work has produced fixed output: tables, figures and reports that say, "here are the results." Shiny is interesting because it lets you move one step further and build a small tool that responds to user input. Instead of preparing one answer, you can let somebody explore several scenarios for themselves.

That is useful more often than people think. Sometimes the same calculation needs to be repeated with different settings. Sometimes a result is easier to understand when you can move a slider or choose a subgroup. And sometimes the most practical way to share an analysis is not as a static report, but as a simple app.

Shiny is the package in `R` that makes that possible. It works well together with RStudio, and once the `shiny` package is installed you can run apps locally on your own computer or host them elsewhere, for example using Posit Cloud.

```{r installation, eval=FALSE}
install.packages("shiny")
```

## Creating an app

There are several ways to create a Shiny application. The easiest way is to use the menus in RStudio. Click `File > New File > Shiny Web App`. You will then be asked to choose an application name and a directory for the app.

Nowadays, the usual process is to create a file called `app.R` that contains both the user interface and the back end. In older Shiny apps, these two parts were often stored in separate files called `ui.R` and `server.R`. You can still encounter that structure, but in this course we will use the single-file `app.R` approach.

When you are more experienced you can also start building an app by creating an `app.R` file in an empty directory yourself. You can quickly enter the boilerplate code typing `shinyapp` and pressing {{< kbd Shift-Tab >}}.

## Parts of an app: user interface and back end

If you inspect the code of a simple Shiny app, you will see that it consists of two main parts:

- a **user interface** (`ui`), which controls what the user sees
- a **server** function, which controls what the app does

```{r uiserver, eval=FALSE}
library(shiny)

ui <- fluidPage(
  # code to define the user interface goes here
)

server <- function(input, output, session) {
  # code to define the back end goes here
}

shinyApp(ui, server)
```

In other words, the user interface is the visible part of the app, while the server contains the logic behind it.

## Starting and stopping the application

When you want to run the app, you can click the `Run App` button in the toolbar of the source pane. You can also use the shortcut {{< kbd Ctrl-Shift-Enter >}}. As another option, you can use the `runApp()` function and provide the directory of the application as an argument.

To stop the application, you can press the stop sign in the toolbar of the console pane, press {{< kbd ESC >}} when the console pane is selected, or simply close the application window.

## Input controls

We want our application to receive input from a user and show results based on that input. These things are done using **controls** placed in the user interface.

Let us first look at input controls and change the `ui` to look like this:

```{r inputcontrol, eval=FALSE}
ui <- fluidPage(
  sliderInput("n",
              "Enter a number:",
              min = 1,
              max = 10,
              value = 2)
)

```

Here `ui` uses `fluidPage()`. This is a **layout function**, meaning that it controls the overall layout of the app. A fluid page layout structures elements in rows and columns and is often a good starting point for a simple app.

Inside the page we place a `sliderInput()`, which allows the user to select a number between 1 and 10.

The most important arguments are:

- the `inputId`, which is the name we use in the server code
- the `label`, which tells the user what the control is for
- the `min`, `max` and `value`, which define the range and default setting

::: callout-note
### What makes Shiny special

The control itself does not yet do anything interesting. The important point is that once the app is running, the selected value becomes available to the server as `input$n`.
:::

## The server function

The app does not do anything useful yet because the `server` function is still empty. If we want to add functionality, we need to define outputs inside that function. Let us simulate the total from a number of dice throws.

```{r dice, eval=FALSE}
server <- function(input, output, session) {
  output$total <- renderText({
    sum(ceiling(runif(input$n, 0, 6)))
  })
}

```

Here we define an output called `total`. This is done using a **render function**, which tells Shiny how the result should be computed and displayed. The `renderText()` function is used when the output can be shown as text.

Inside that code we refer to `input$n`, which is the value currently selected with the slider.

## Output controls

When we actually want to show the output, we need to add a matching output control to the user interface. This can be done as follows:

```{r addsoutput, eval=FALSE}
ui <- fluidPage(
  sliderInput("n",
              "Enter a number:",
              min = 1,
              max = 10,
              value = 2),
  textOutput("total")

)

```

The key idea is that outputs are defined in the server and displayed in the user interface. The name has to match: `output$total` in the server is paired with `textOutput("total")` in the UI.

## What to remember

Every simple Shiny app has the same basic pattern:

1. create a user interface
2. define inputs in the UI
3. refer to those inputs in the server
4. create outputs in the server
5. place output controls in the UI

In the practical, you will build exactly this pattern step by step.

## Check yourself

Before moving on, make sure you can answer the following:

1. What is the difference between the `ui` and the `server` in a Shiny app?
2. What does `input$n` mean?
3. Why do we need both `renderText()` and `textOutput()`?
4. What changes when the app becomes reactive rather than static?
