## Shiny: Questions {.unnumbered}

The goal of this practical is to create a Shiny application that visualises birthweight percentiles using the [Hoftiezer Birthweight curves](https://www.perined.nl/geboortegewichtcurven). These curves are widely used to determine the birthweight percentile of an infant and to assess whether a child may be small for gestational age (SGA).

By completing the exercises, you will:

- use an external `R` function inside a Shiny app
- create reactive user inputs and outputs
- calculate percentiles and reference curves for specified gestational ages
- integrate these calculations into a clear, user-friendly interface

Open RStudio to do the practicals.

::: callout-tip
Run the app after every major step. In Shiny work, frequent testing is one of the fastest ways to spot mistakes.
:::

::: panel-tabset
### Question 1

Download the `hoftiezer.R` file from Canvas to get started. This file contains two functions you can use: `phoftiezer` and `qhoftiezer`. The first calculates Hoftiezer birthweight percentiles given birthweight (in grams), gestational age (in days) and sex. The second does the reverse and calculates birthweight given a percentile, gestational age and sex. You do not need to know every detail of these functions, but you do need to know how to call them correctly.

In this question, you will practice using the functions outside of Shiny first. These calculations will later be incorporated into the app.

-   Using this function, calculate the birthweight percentile of a boy with a gestational age of 200 days who weighs 1500 grams. Also calculate the percentile for this same weight for all integer gestational ages between 210 and 220 days.

-   Using the other function, calculate the median birthweight of a girl with a gestational age of 200 days. Also calculate the P05 (5th percentile) for girls for all integer gestational ages between 210 and 220 days.
:::

::: panel-tabset
### Question 2

Now we are going to make the Shiny app itself. Although there is a wizard to create Shiny apps, here we will start from a basic `R` file so that the structure is fully visible.

Create a new directory for your app. Then create a new `R` file, save it in that directory, and give it the name `app.R`. Copy and paste the following code into that file:

```{r, eval=FALSE}
library(shiny)

ui <- fluidPage(
    # Application title
    titlePanel("Birthweight percentiles")
)

server <- function(input, output) {
# nothing here yet
}

# Run the application
shinyApp(ui = ui, server = server)
```

If you save the file, a green arrow should appear in the source toolbar in RStudio. When you click it, the app will run. At this point it will not do very much, but it should display its title.

::: callout-note
## Milestone

At this point, you should have a working but very simple Shiny app that launches successfully.
:::
:::

::: panel-tabset
### Question 3

We want our app to be able to calculate birthweight percentiles. Therefore it needs access to the `qhoftiezer` and `phoftiezer` functions. One way would be to copy the contents of `hoftiezer.R` into `app.R`, but a cleaner approach is to source that file.

Save the `hoftiezer.R` file in the same directory as your app. Then include the line `source("hoftiezer.R")` in `app.R`, just before the definition of `ui` is a good place. To test whether the app can access the function, use one of the functions directly in the interface. The start of your file should now look like this:

```{r, eval=FALSE}
library(shiny)

source("hoftiezer.R")
ui <- fluidPage(
  # Application title
  titlePanel("Birthweight percentiles"),
  qhoftiezer(p = 0.5, ga = 220, sex = "female")
)

# ...
```

Render and run the app again. If you see a number appear in the interface, the file was sourced successfully.
:::

::: panel-tabset
### Question 4

We will now modify the layout of the app so that user input appears on the left and results appear on the right. In Shiny this can be done with a `sidebarLayout()` inside `fluidPage()`, together with a `sidebarPanel()` and a `mainPanel()`.

We start with only one control, for birthweight, and we will add the other inputs later. Modify the `ui` so it looks like this:

```{r, eval=FALSE}
ui <- fluidPage(
    # Application title
    titlePanel("Birthweight percentiles"),
    sidebarLayout(
        sidebarPanel(
          numericInput("birthweight",
                        "Birthweight in grams:",
                        min = 100,
                        max = 6000,
                        value = 1000)
    ),
        # Show percentile
        mainPanel(
        phoftiezer(q = 0.5, ga = 220, sex='female')
        )
    )
)

# ...
```

Within `numericInput()`, the first parameter specifies the `inputId`, the name we will use to refer to the value entered by the user. The second parameter specifies the label that is displayed to the user. The other arguments specify the minimum and maximum value and the default value.

::: callout-note
## Milestone

At this point, your app should have a proper layout with a sidebar and a main panel, even if the result is still static.
:::
:::

::: panel-tabset
### Question 5

You have probably noticed that in its present form the app doesn't take any action based on the birth weight entered by the user. To address this, it's essential to conduct the required calculations within our server function. Update the server function to resemble the following:

```{r, eval=FALSE}
server <- function(input, output) {
  output$percentile <- renderText({
    phoftiezer(q = input$birthweight,
               ga = 220,
               sex = 'female')
  })

```

With the assignment `output$percentile <- renderText(...)` you define a Shiny output with the name `percentile`. The right-hand side is a render function. In this case, the result will be shown as text. Inside `phoftiezer()` we now use `input$birthweight`, which means the calculation depends on the current input value.

To display the result from `output$percentile`, replace the direct function call in the main panel with `textOutput("percentile")`.

Verify if the app behaves as you expect.

::: callout-note
## Milestone

At this point, your app should react when the user changes the birthweight.
:::
:::

::: panel-tabset
### Question 6

Additionally, we would like to specify the infant's sex and gestational age. To do this, add two more controls to the `sidebarPanel()`: use a `sliderInput()` for gestational age and a `selectInput()` for sex. Use the documentation if you need to check how these functions work.

Do not forget that input controls in the `sidebarPanel()` need to be separated by commas. Then modify the `server` function so it uses the values from these controls.

::: callout-note
## Milestone

At this point, your app should respond to three user inputs:

- birthweight
- gestational age
- sex
:::
:::

::: panel-tabset
### Question 7

To make the app more informative, we will include a plot. Add the line `plotOutput("distPlot")` to the `mainPanel()`. Then define `output$distPlot` inside the server.

Inside that plot code:

- define a sequence of gestational ages from 161 to 294
- use `qhoftiezer()` to calculate the median birthweight curve
- do the same for the 5th and 95th percentile curves
- plot these curves
- add the current user input as a point on the plot

You can use base plotting functions such as `plot()`, `lines()` and `points()`.

::: callout-note
## Final check

Before you finish this practical:

- run the app once more from top to bottom
- check that the app starts without errors
- verify that the text output changes when the controls change
- verify that the plotted point moves when the controls change
:::
:::
