---
title: "Analysis of Intellectual Disabilities"
author: "Sten Willemsen"
date: "2024-06-06"
format:
  html:
    code-fold: true
---

## Setup

```{r setup}
dat <- read.csv("Data/R_data2.csv")
```

## Introduction

In this report, we analyse data on children with intellectual disabilities. The report combines code, results and explanation in one document so that the full workflow remains transparent and easy to update.

## Data

We perform the following data transformation steps:

* We define the variable `pregnancy_length` by combining the pregnancy length in weeks and days.

* We define the variable `BMI_cat` by dividing the variable BMI into categories: <18.5 ("Underweight"), 18.5 - 24.9 ("Healthy weight"), 25 - 29.9 ("Overweight"), and >30 ("Obesity").

* We log transform `homocysteine` and `vitaminB12`.

* We transform the categorical variables to factors.

* We remove the original variables `pregnancy_length_weeks`, `pregnancy_length_days`, `BMI`, `homocysteine` and `vitaminB12` from the data set.

```{r transformations}
#| echo: false
#| message: false
dat$pregnancy_length <- dat$pregnancy_length_weeks * 7 +
  dat$pregnancy_length_days

dat$BMI_cat <- cut(dat$BMI,
                   breaks = c(-Inf, 18.5, 24.9, 29.9, Inf),
                   labels = c("Underweight", "Healthy weight", "Overweight", "Obesity"))

dat$log_homocysteine <- log(dat$homocysteine)
dat$log_vitaminB12 <- log(dat$vitaminB12)

for (i in 1:length(names(dat))) {
  if (class(dat[[i]]) == "character") {
    dat[[i]] <- as.factor(dat[[i]])
  }
}

dat$Status <- factor(dat$Status,
                     levels = c("normal brain development", "intellectual disability"))

dat <- dat[, !(names(dat) %in% c("pregnancy_length_weeks",
                                 "pregnancy_length_days",
                                 "BMI",
                                 "homocysteine",
                                 "vitaminB12"))]
```

These transformations make the data more suitable for the analyses below and make the coding of key variables more explicit.

## Analysis and Results

We first show descriptives of all variables in the data set and then perform both an unadjusted and an adjusted analysis.

### Descriptives

```{r descriptives}
for (i in 1:length(names(dat))) {
  if (class(dat[[i]]) == "numeric") {
    print(paste("Mean of", names(dat)[i], ":", mean(dat[[i]], na.rm = TRUE)))
    print(paste("Standard deviation of", names(dat)[i], ":", sd(dat[[i]], na.rm = TRUE)))
  } else if (class(dat[[i]]) == "factor") {
    print(paste("Frequency of", names(dat)[i], ":"))
    print(table(dat[[i]]))
  }
}
```

For the continuous variables, we also make histograms.

```{r}
for (i in 1:length(names(dat))) {
  if (class(dat[[i]]) == "numeric") {
    hist(dat[[i]], main = paste("Histogram of", names(dat)[i]))
  }
}
```

### Unadjusted Analysis

We compare the mean of the logarithm of Vitamin B12 for the two levels of `Status` (normal brain development or intellectual disability).

```{r}
t.test(log_vitaminB12 ~ Status, data = dat)
```

## Adjusted Analysis

We now perform logistic regression analysis to investigate the association between `Status` and log `Vitamin B12` while adjusting for `medication`, `smoking` and `alcohol`.

```{r}
glm1_adjusted <- glm(Status ~ log_vitaminB12 + medication + smoking + alcohol,
                     data = dat,
                     family = binomial)

summary(glm1_adjusted)
coef(glm1_adjusted)
confint(glm1_adjusted)
```

## Conclusion and Discussion

**Main points:**

* In the unadjusted analysis, we did not find evidence that the mean of the logarithm of Vitamin B12 differs significantly between the two levels of `Status`.

* In the adjusted analysis, we found that log `Vitamin B12` is not significantly associated with `Status` after adjusting for `medication`, `smoking` and `alcohol`.
